---
title: "Epidemic Curve & Herd Immunity Threshold Analysis"
author: "Epidemiology Unit"
date: today
format:
  docx:
    toc: true
    number-sections: true
    fig-width: 7.5
    fig-height: 4.5
    fig-dpi: 180
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(scales)
library(lubridate)
library(ggrepel)
library(knitr)
```

## Load and Inspect Data

```{r load}
df <- read_csv("epidemic_data.csv",
               col_types = cols(onset_date = col_date()))
glimpse(df)
df |> slice_head(n = 6) |> kable(caption = "First 6 rows of dataset")
```

```{r summary-stats}
df |>
  group_by(disease, r0, hit_pct) |>
  summarise(
    cases      = n(),
    pct_unvacc = round(mean(vaccinated == "No") * 100, 1),
    pct_severe = round(mean(severity   == "Severe") * 100, 1),
    .groups = "drop"
  ) |>
  arrange(desc(r0)) |>
  kable(caption = "Outbreak summary by disease")
```

Measles and Pertussis share R0 = 15 and the highest HIT (93.3%). Unvaccinated cases exceed 65% across all diseases.

---

## Epidemic Curves

### Overall Weekly Incidence (Measles)

```{r epi-curve-overall}
#| fig-cap: "Weekly measles case counts. Peak around week 6-7 confirms propagated-source transmission."

measles <- df |>
  filter(disease == "Measles") |>
  mutate(week = floor_date(onset_date, "week")) |>
  count(week, name = "cases")

pk <- measles |> filter(cases == max(cases))

ggplot(measles, aes(week, cases)) +
  geom_col(fill = "#C0392B", colour = "white",
           width = 6, alpha = 0.88) +
  geom_smooth(method = "loess", se = FALSE,
              colour = "#2C3E50", linewidth = 0.9,
              linetype = "dashed") +
  annotate("label", x = pk$week, y = pk$cases + 8,
           label = paste0("Peak: ", pk$cases, " cases"),
           size = 3, colour = "#C0392B",
           fill = "white", label.size = 0.3) +
  scale_x_date(date_labels = "%d %b",
               date_breaks = "2 weeks") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .15))) +
  labs(title = "Measles Epidemic Curve — Weekly Cases",
       x = "Week of Onset", y = "Cases") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )
```

### Age-Stratified Curve

```{r epi-curve-age}
#| fig-cap: "Children 5-14 and young adults 15-29 bear the greatest burden."

df |>
  filter(disease == "Measles") |>
  mutate(week = floor_date(onset_date, "week")) |>
  count(week, age_group) |>
  ggplot(aes(week, n, fill = age_group)) +
  geom_col(colour = "white", width = 6, alpha = 0.90) +
  scale_fill_manual(
    values = c("0-4"  = "#E74C3C",
               "5-14" = "#E67E22",
               "15-29"= "#3498DB",
               "30-49"= "#27AE60",
               "50+"  = "#8E44AD"),
    name = "Age group") +
  scale_x_date(date_labels = "%d %b",
               date_breaks = "2 weeks") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .1))) +
  labs(title = "Measles: Age-Stratified Epidemic Curve",
       x = "Week of Onset", y = "Cases") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )
```

### Multi-Disease Incidence Comparison

```{r epi-curve-multi}
#| fig-cap: "High-R0 diseases peak earlier and steeper than low-R0 ones."
#| fig-height: 6

df |>
  mutate(week = floor_date(onset_date, "week")) |>
  count(disease, week) |>
  ggplot(aes(week, n, fill = disease)) +
  geom_col(width = 5, alpha = 0.85, colour = NA) +
  facet_wrap(~ disease, ncol = 2, scales = "free_y") +
  scale_x_date(date_labels = "%b",
               date_breaks = "1 month") +
  scale_y_continuous(
    expand = expansion(mult = c(0, .1))) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "Epidemic Curves by Disease (Weekly)",
       x = "Month", y = "Cases") +
  theme_minimal(base_size = 9) +
  theme(
    plot.title       = element_text(face = "bold"),
    strip.text       = element_text(face = "bold"),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

High-R0 diseases (Measles, Pertussis) produce sharp narrow peaks. Influenza and Ebola spread more slowly with flatter curves.

---

## Herd Immunity Threshold

$$HIT = 1 - \frac{1}{R_0}$$

When population immunity >= HIT, each case generates fewer than one secondary infection and the outbreak declines.

```{r hit-reference}
df |>
  distinct(disease, r0, hit_pct) |>
  arrange(desc(r0)) |>
  rename(Disease = disease, R0 = r0, `HIT (%)` = hit_pct) |>
  kable(caption = "Herd immunity thresholds by disease")
```

### HIT Bar Chart

```{r hit-bar}
#| fig-cap: "Measles and Pertussis require >93% coverage."

df |>
  distinct(disease, r0, hit_pct) |>
  ggplot(aes(reorder(disease, hit_pct),
             hit_pct, fill = hit_pct)) +
  geom_col(width = 0.65, alpha = 0.92,
           show.legend = FALSE) +
  geom_text(aes(label = paste0(hit_pct, "%")),
            hjust = -0.15, size = 3.2,
            fontface = "bold") +
  geom_hline(yintercept = 70,
             linetype = "dashed",
             colour = "#E67E22", linewidth = 0.7) +
  annotate("text", x = 0.7, y = 71.5,
           label = "70% line",
           colour = "#E67E22", size = 2.8) +
  scale_fill_gradient(low = "#F8C471",
                      high = "#922B21") +
  scale_y_continuous(limits = c(0, 107),
    labels = label_percent(scale = 1)) +
  coord_flip() +
  labs(title = "Herd Immunity Threshold by Disease",
       x = NULL, y = "HIT (%)") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title         = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

### Continuous HIT-R0 Curve

```{r hit-curve}
#| fig-cap: "Non-linear growth: small R0 increases demand disproportionate immunity gains."

pts <- df |> distinct(disease, r0, hit_pct)

tibble(R0  = seq(1.01, 20, 0.05),
       HIT = (1 - 1/R0) * 100) |>
  ggplot(aes(R0, HIT)) +
  geom_line(colour = "#2980B9", linewidth = 1.3) +
  geom_area(fill = "#AED6F1", alpha = 0.3) +
  geom_point(data = pts,
    aes(x = r0, y = hit_pct, colour = disease),
    size = 3, show.legend = TRUE) +
  geom_label_repel(
    data = pts,
    aes(x = r0, y = hit_pct,
        label = disease, colour = disease),
    size = 2.5, show.legend = FALSE,
    label.size = 0.2, max.overlaps = 10) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  scale_y_continuous(
    labels = label_percent(scale = 1),
    breaks = seq(0, 100, 10)) +
  labs(title = "HIT as a Function of R0",
       x = "Basic Reproduction Number (R0)",
       y = "Herd Immunity Threshold (%)",
       colour = "Disease") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

---

## Vaccination & Severity Analysis

```{r vacc-analysis}
#| fig-cap: "Unvaccinated individuals dominate case counts across all diseases."

df |>
  count(disease, vaccinated) |>
  group_by(disease) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(reorder(disease, pct),
             pct, fill = vaccinated)) +
  geom_col(width = 0.65, alpha = 0.90) +
  geom_hline(yintercept = 50,
             linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("Yes" = "#27AE60",
               "No"  = "#C0392B"),
    name = "Vaccinated") +
  scale_y_continuous(
    labels = label_percent(scale = 1)) +
  coord_flip() +
  labs(title = "Case Distribution by Vaccination Status",
       x = NULL, y = "% of Cases") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title         = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

```{r severity-analysis}
#| fig-cap: "Severe cases are consistently higher among unvaccinated patients."

df |>
  count(disease, vaccinated, severity) |>
  group_by(disease, vaccinated) |>
  mutate(pct = n / sum(n) * 100) |>
  filter(severity == "Severe") |>
  ggplot(aes(disease, pct, fill = vaccinated)) +
  geom_col(position = "dodge",
           width = 0.6, alpha = 0.90) +
  scale_fill_manual(
    values = c("Yes" = "#27AE60",
               "No"  = "#C0392B"),
    name = "Vaccinated") +
  scale_y_continuous(
    labels = label_percent(scale = 1)) +
  labs(title = "% Severe Cases by Disease & Vaccination Status",
       x = NULL, y = "% Severe") +
  theme_minimal(base_size = 10) +
  theme(
    plot.title       = element_text(face = "bold"),
    axis.text.x      = element_text(angle = 40, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

Unvaccinated patients show higher severe-case proportions across all diseases. The severity gap is widest for Measles and Pertussis where vaccine-derived protection matters most.

---

## Key Findings

- **Epidemic shape:** High-R0 diseases produce narrow tall peaks; low-R0 diseases produce flat prolonged curves.
- **HIT non-linearity:** R0 = 15 requires 93.3% immunity vs 60% for R0 = 2.5 — a small absolute gap with large operational consequences.
- **Vaccination gap:** Over 65% of cases across all diseases were unvaccinated — the primary modifiable risk factor.
- **Severity:** Unvaccinated individuals carry a disproportionate share of severe outcomes, reinforcing the individual and population benefit of vaccination.
